---
title: "Test for Equilibrium"
author: "Group 1: Gabriel B, Innocent, Marco, Thales"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Data

We loaded the data and identified which alleles are fixed.

```{r}

data <- read.table("BIO641.txt", header = TRUE, sep = "\t")

constante <- c()
for (i in 1:ncol(data)) {
  if (length(unique(data[, i])) == 1) {
    constante <- c(constante, i)
  }
}

data_cleaned <- data[, -constante]
```

## Count

We counted the homozygotes and heterozygotes for each marker in the filtered dataset.

```{r}
count_0 <- integer(ncol(data_cleaned))
count_1 <- integer(ncol(data_cleaned))
count_2 <- integer(ncol(data_cleaned))

for (i in 1:ncol(data_cleaned)) {
  count_0[i] <- sum(data_cleaned[, i] == 0)
  count_1[i] <- sum(data_cleaned[, i] == 1)
  count_2[i] <- sum(data_cleaned[, i] == 2)
}

result <- data.frame(
  Column = colnames(data_cleaned),
  Count_0 = as.numeric(count_0),
  Count_1 = as.numeric(count_1),
  Count_2 = as.numeric(count_2)
)
```

## Chi-Square Test for Equilibrium

We calculated the expected value based on theoretical equilibrium and then compared it with the observed values to compute the chi-square value.

```{r}
qui_values <- numeric(nrow(result))
p_values <- numeric(nrow(result))

for (i in 1:nrow(result)) {
  q <- ((result[i, "Count_0"] + 0.5 * result[i, "Count_1"]) / 
        (result[i, "Count_0"] + result[i, "Count_1"] + result[i, "Count_2"]))
  p <- 1 - q
  total <- result[i, "Count_0"] + result[i, "Count_1"] + result[i, "Count_2"]
  
  expectedAA <- (p^2) * total
  expectedAa <- (2 * p * q) * total
  expectedaa <- (q^2) * total
  
  qui <- (((result[i, "Count_2"] - expectedAA)^2 / expectedAA) + 
          ((result[i, "Count_1"] - expectedAa)^2 / expectedAa) + 
          ((result[i, "Count_0"] - expectedaa)^2 / expectedaa))
  
  qui_values[i] <- qui
  
  if (qchisq(0.95, df = 1) > qui_values[i]) {
    p_values[i] <- "eq"
  } else {
    p_values[i] <- "neq"
  }
}
```

## Correction for Multiple Tests

We applied the Bonferroni correction for multiple tests and evaluated whether the results still indicate equilibrium.

```{r}
astar <- 0.05 / nrow(result)
f <- 1 - astar
p_values_corrected <- character(nrow(result))

for (i in 1:length(qui_values)) {
  if (qchisq(f, df = 1) > qui_values[i]) {
    p_values_corrected[i] <- "eq"
  } else {
    p_values_corrected[i] <- "neq"
  }
}
```

## Final Results

The results are presented in a final table with the markers, equilibrium without correction, and after correction.

```{r}
dataframe <- data.frame(
  Markers = result$Column,
  Equilibrium = p_values,
  `Equilibrium Corrected` = p_values_corrected
)

dataframe
```


